---
title: "Logistic Regression"
teaching: 60
exercises: 60
source: Rmd
---

::::::::::::::::::::::::::::::::::::::: objectives

- To be able to construct regression models for a binary outcome
- To be able to calculate predicted variables and residuals
- To be able to present model outcomes using Broom

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I identify factors for antibiotic resistance?
- How can I check the validity of model?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Content

-   Exploratory Data Analysis
-   Model Creation and Estimation
-   Reporting the Logistic Regression Results

## Data

```{r libraries, message=FALSE, warning=FALSE}
# We will need these libraries and this data later.
library(aod)
library(broom)
library(ggplot2)
library(lubridate)
library(odds.n.ends)
library(tidyverse)

amrData <- read.csv("data/dig_health_hub_amr.csv")
```
The data used in this episode was provided by Simon Thelwall from the UKHSA.
It has been created to represent the sort of data that might be obtained from the Second Generation Surveillance System (SGSS).
The data has 100,000 rows of 12 variables.

## Exploratory Data Analysis

We can preview our data by using 'head':

```{r}
head(amrData)
```
We can also request information about variable names and data types:
```{r}
sapply(amrData, class)
```

We can see that dates are currently stored as the char data type. We also do not know the age of the participant when the sample was taken.

```{r}
# Calculate age (in years) as of their last birthday and add as an additional variable to our data.

amrData <- amrData %>%
  mutate(
    age_years_sd = (dob %--% spec_date) %/% years(1)
  )
```

We can also convert 'spec_date', the date the specimen was taken from text to a date:
```{r}
# Convert char to date and store as additional variable
amrData <- amrData %>%
  mutate(
    spec_date_YMD = as.Date(amrData$spec_date)
  )
```
We can use a histogram to explore the age distribution of the participants:

```{r}
# histogram of age
ageHisto <- amrData %>%
  ggplot(aes(x = age_years_sd)) +
  geom_histogram(bins = 10, color = "white") +
  theme_minimal(base_size = 14, base_family = "sans") +
  labs(x = "Age when specimen taken (years)", y = "Frequency")
ageHisto
```
We can also look at where the specimens were processed:
```{r}
xtabs(~region, data = amrData)
```

and for which organism:
```{r}
xtabs(~organism, data = amrData)
```

In addition, we can use cross-tabulation to identify if the specimen indicated resistance to one or more of Coamoxiclav, Gentamicin and Ciprofloxacin for the participants:

```{r}
xtabs(~ coamox + cipro + gentam, data = amrData)
```
We can see from our table that only 227 participants indicated reistance to Coamoxiclav, Gentamicin and Ciprofloxacin.
Coamoxiclav appears to have the highest individual indication of resistance. We will explore indicators to Coamoxiclav first.

##  Model Creation and Estimation

As the dependent variable we want to explore is binary (0,1), we will use a binomial generalised linear model.


```{r}
coamox_logit <- glm(coamox ~ age_years_sd + sex_male, data = amrData, family = "binomial")

summary(coamox_logit)
```
In this initial model focusing on demographic indicators, age at time of sample being taken and whether or not the participant is male both seem to be statistically significant.

age_years_sd: For every unit increase in age_years_sd the log-odds of Coamoxiclav resistance increase by 0.0404280.

sex_male: The difference in the log-odds of Coamoxiclav resistance between males and non-males is 0.0540494.

Older participants and male participants have higher log-odds of Coamoxicalv resistance.

We can check to see that our indicators sex_male and age_years_sd are independent:
```{r}
# check VIF for no perfect multicollinearity assumption
car::vif(coamox4_logit)
```
We can also check the linearity of the variable age_years_sd

```{r}
# make a variable of the logit of the predicted values
logit.use <- log(coamox_logit$fitted.values / (1 - coamox_logit$fitted.values))

# make a small data frame with the logit variable and the age predictor
linearity.data <- data.frame(logit.use, age = coamox_logit$model$age_years_sd)

# create a plot with linear and actual relationships shown
linearPlot <- linearity.data %>%
  ggplot(aes(x = age, y = logit.use)) +
  geom_point(aes(size = "Observation"), color = "blue", alpha = .6) +
  geom_smooth(se = FALSE, aes(color = "Loess curve")) +
  geom_smooth(method = lm, se = FALSE, aes(color = "linear")) +
  theme_minimal(base_size = 14, base_family = "serif") +
  labs(x = "Age in years on sample date", y = "Log-odds of coamox resistance predicted probability") +
  scale_color_manual(name = "Type of fit line", values = c("red", "black")) +
  scale_size_manual(values = 1.5, name = "")

linearPlot
```

::::::::::::::::::::::::::::::::::::::  challenge

### Challenge 1

Update the model coamox_logit to include had_surgery_past_yr as an independent variable.
What is the log-odds reported and is it statistically significant?

:::::::::::::::  solution

### Challenge 1

You may choose to create a new glm:

```{r}
coamox_surg_logit <- glm(coamox ~ age_years_sd + sex_male + had_surgery_past_yr, data = amrData, family = "binomial")

summary(coamox_surg_logit)
```
had_surgery_past_yr: The difference in the log-odds of Coamoxiclav resistance between those who have had surgery in the past year
and those who have not is 0.1802123. It is statistically significant.

You may also want to check for culticollinearity:

```{r}
car::vif(coamox_surg_logit))
```
As the value of GVIF is lower than 4, it suggests that the assumption of independnce between the variables is held.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

### Incorporating a multi-level factor

So far we have considered binary and continuos variable indicators in our model(s).
Our data set also contains some categorical variables: region, ethinicity and imd.

We are going to incorporate region into our model:

```{r}
# To include reporting for all regions I will include the 0 flag
coamox_region_logit <- glm(coamox ~ 0 + age_years_sd + sex_male + region, data = amrData, family = "binomial")
summary(coamox_region_logit)
```
It initially seems that all regions are significant and have a negative log-odds of Coamoxiclav resistance.


We can use the Wald test to explore the overall effect of region. We will use the wald.test function of the aod library. The order in which the coefficients are given in the table of coefficients is the same as the order of the terms in the model. This is important because the wald.test function refers to the coefficients by their order in the model. We use the wald.test function. b supplies the coefficients, while Sigma supplies the variance covariance matrix of the error terms, finally Terms tells R which terms in the model are to be tested, in this case, terms 3 to 11, are the three terms for the different regions.

```{r}
# model with just the regions added
wald.test(b = coef(coamox_region_logit), Sigma = vcov(coamox_region_logit), Terms = 3:11)
```

The chi-squared test statistic of 12815.9, with nine degrees of freedom is associated with a p-value of 0.00 indicating that the overall effect of region is statistically significant.

::::::::::::::::::::::::::::::::::::::  challenge

### Challenge 2

Update the model you created in Challenge 1 to include either ethnicity or imd as an independent variable.
What is the log-odds reported and is it statistically significant?

:::::::::::::::  solution

### Challenge 2

If you chose to incorporate imd, this would first need to be converted to a factor

```{r}
amrData$imd <- factor(amrData$imd)

coamox_surg_imd_logit <- glm(
  coamox ~ age_years_sd + sex_male + had_surgery_past_yr + imd,
  data = amrData,
  family = "binomial"
)

summary(coamox_surg_imd_logit)
```
As we did not include the 0 flag IMD quintile 1 has been used a reference for the other Quintiles.
Quintiles 2 to 5 are not significantly significant.
We may want to test that the coefficient for imd=1 is equal to the coefficient for imd=2.
The first line of code below creates a vector l that defines the test we want to perform. In this case, we want to test the difference (subtraction) of the terms for imd=1  and imd=2  (i.e., the 4th and 5th terms in the model). To contrast these two terms, we multiply one of them by 1, and the other by -1. The other terms in the model are not involved in the test, so they are multiplied by 0. The second line of code below uses L=l to tell R that we wish to base the test on the vector l (rather than using the Terms option as we did above).

```{r}
l <- cbind(0, 0, 0, 1, -1, 0, 0, 0)
wald.test(b = coef(coamox_surg_imd_logit), Sigma = vcov(coamox_surg_imd_logit), L = l)
```
The chi-squared test statistic of 33.9, with 1 degree of freedom is associated with a p-value of 5.7e-09 indicating that the overall effect of region is statistically significant.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Reporting the Logistic Regression Results

For our model coamox_region_logit there are various things that we can report, and different functions and packages that can be used.

We can report the CIs using profiled log-likelihood or using standard errors for our model.
```{r}
# CIs using log-likelihood
confint(coamox_region_logit)

## CIs using standard errors
confint.default(coamox_region_logit)
```

Alternatively we may be interested in the odds-ratio:
```{r}
## odds ratios only
exp(coef(coamox_region_logit))

# including confidence intervals
exp(cbind(OR = coef(coamox_region_logit), confint(coamox_region_logit)))
```
These can also be presented using broom:
```{r}
# as Log-Odds
tidy(coamox_region_logit)

# as ORs
tidy(coamox_region_logit, exp = TRUE)
```

The odd.n.ends package provides a wide range of reporting tools

```{r}
# Odds ratios and confidence intervals
coamox_region_logitOR <- odds.n.ends::odds.n.ends(coamox_region_logit)
coamox_region_logitCI <- coamox_region_logitOR$`Predictor odds ratios and 95% CI`
coamox_region_logitCI

# model fit
modfit <- coamox_region_logitOR$`Count R-squared (model fit): percent correctly predicted`
modfit

# Other model statistics
odds.n.ends::odds.n.ends(coamox_region_logit)
```
As you can see we have a range of probabilities, odds ratios and log-odds. We need to be aware of which we are referring to.
One way is to look at the range of the estimates. Probabilities always have a range from zero to 1. Logit units generally range from about -4 to +4, with zero meaning an equal probability of no event or the event outcome occurring. Odds ratios can range from very small (but positive) numbers to very large positive numbers.

These odds ratios versions of the estimates are more easily interpretable than logit scores. Odds ratios of less than one means that an increase in that predictor makes the outcome less likely to occur, and an odds ratio of greater than one means that an increase in that predictor makes the outcome more likely to occur.

::::::::::::::::::::::::::::::::::::::: discussion

## Discussion

* What other indicators could we have included in our model?
* What question(s) would they help us to answer?

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 3

We have explored potential indicators of Coamoxiclav resistance.

Explore potential indicators of either Gentamicin or Ciprofloxacin resistance.

Present your analysis steps and findings in a Quarto report.

::::::::::::::::::::::::::::::::::::::::::::::::::
